---
title: "Getting help"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "show"
---

```{css}
# This chunk contains 'CSS' code, which alters the appearance of
# the html output
h1 {background: brown;color: white;padding-left: 7px; font-size:large}
h2 {background: grey;
color: white;
padding-left: 7px;
padding-top:5px;
padding-bottom:5px;
font-size:medium}
h3 {color: brown;font-size:medium}
code {color: chocolate;}
```



## The Programmer's Categorical Imperative  

***Whenever possible, copy pre-existing code - then modify for your own purposes!***

## (1) To get help on a specific function the first option usually is to use the 'Help' tab in RStudio
The Help tab is usually in the bottom-right window in RStudio.  

## (2) Help is also available through code:  
The help file will open in te RStudio Help tab (usually in the bottom-right window).  
```{r}
help(sd)
```

A question mark before the function is equivalent to the `help()` function:  
```{r}
?median
```


Typing two question marks in front of a keyword will search for that word anywhere in the help files: 
```{r}
??variance
```


## (3) Sometimes the same function name is used by different packages. To get help on a function from a specific package, use:  
```{r}
help(sd, package="stats")
```


## (4) For most functions, there are code examples:  
```{r results="hide"}
example(persp)
```


## (5) To see a list of help files for _all_ functions from a specific package, use:
```{r}
help(package="stats")
```


## (6) For some (not all) packages, so-called 'vignettes' are available 
These are articles that explain packages with code examples that show how to work through a typical problem. If no vignette is available from within R, go to the package's page on CRAN or Bioconductor respectively.
```{r}
vignette("magrittr")
```


## (7) All programmers spend a lot of time searching for help on Google  

Mostly looking for search hits on the stackoverflow web site:

* [https://stackoverflow.com/documentation/r](https://stackoverflow.com/documentation/r)
* [https://stackoverflow.com/documentation/r/topic](https://stackoverflow.com/documentation/r/topic)

## (8) To stay up-to-date on R consider subscribing to r-bloggers:

* https://www.r-bloggers.com/  
* https://feeds.feedburner.com/RBloggers  
